Unbiased pangenome graphs

### 2022-02-14 manuscript/main.tex
mention algorithm 1
spell check

In this section, we provide a formal definition of variation graph induction.
We then examine the bounds of a na\"{i}ve implementation of this algorithm.
Finally, we propose compression and partitioning techniques to reduce the space and working memory complexity of the induction process by a large constant factor modulated by the degree of sequence divengence in the input pangenome.
This yields a practical algorithm for variation graph induction that can scale to the largest available pangenomes.

@@

The same encoding can be used to replace the sequence-to-graph mapping $Z$, yielding $\mathcal{Z}$.
Rather than pairs of characters in $Q$ and $\mathcal{N}$, we record runs of matches between them as range matches.
Although in expectation the length of these matches should be strictly less than $k$, due to the interruption of the graph by variation between genomes, this still allows us to reduce the size of $Z$ using runs of matches between $Q$ and $\mathcal{N}$.
Additionally, we store the inverse of $\mathcal{Z}$, which maps ranges from $\mathcal{N} \to Q$, as $\bar{\mathcal{Z}}$.
We use $\mathcal{Z}$ to compact unbranching regions of $\mathcal{V}$ into single nodes, and $\bar{\mathcal{Z}}$ to accelerate our calculation of links in the graph.

@@

To compact $\mathcal{V}$, we traverse $Q$, finding each entry in $\mathcal{Z}$ in turn, recording its start and end in $\mathcal{N}$, which can be understood as a character vector or string containing all the sequence in the nodes of $\mathcal{V}$.
%in a ``graph vector'' $\mathcal{G} = n_1,n_2 \ldots n_{|\mathcal{N}|}$ containing the ordered characters of the graph.
%Any position in $\mathcal{G}$ so marked must be the start or end of a node.
We subsequently use these markings to subdivide $\mathcal{N}$ into a compacted version $\mathcal{N}'$ where compacted node boundaries are marked in a auxilary bitvector $\mathcal{B} : |\mathcal{B}| = |\mathcal{N}|$ where the first character in each compacted node is marked by a $1$ and other characters are marked $0$.
$\mathcal{B}$ allows us to compute compacted node ids using efficient rank operations \citep{Gog_2014}.
%such that $\forall_{i \in 1 \ldots |\mathcal{N}'|} |n_i| \ge 1$.

@@

We have presented a complete model for variation graph induction from sequences and their pairwise alignments.
Here, we describe our specific implementation of this algorithm: \textit{seqwish}.
In general, our approach uses external memory to elaborate the graph, taking advantage of the availability of low-latency storage media, like solid-state drives (SSDs), to maximize the performance of this approach.

@@

We have presented a straightforward algorithm to generate a pangenome graph from a collection of genomes and alignments between them.
By exploiting a simple model of this algorithm, we provide computational bounds that give insight into the complexity of the problem.
We then make this approach practical by applying the concept of \textit{match compression}, which reduces the expected computational complexity by a factor proportional to the diversity of input sequences.
Our experimental results demenstrate that we can apply our method to various collections of sequences and alignments.
It easily scales to some of the largest species pangenome construction problems possible using publicly-available, high-quality genome assemblies.
%Although we have focused on the problem of building pangenomes,
\textit{seqwish} is a generic sequence graph inducer of potentially many uses.
We envision that it can serve as a component in diverse sequence analysis and assembly pipelines, and hope that our thorough description of its core algorithm and functionality will enable its reuse by other researchers.

@@

\section*{Acknowledgements}

>>>>

In this section, we provide a formal definition of variation graph induction.
We then examine the bounds of a na\"{i}ve implementation of this algorithm.
Finally, we propose compression and partitioning techniques to reduce the space and working memory complexity of the induction process by a large constant factor modulated by the degree of sequence divergence in the input pangenome.
This yields a practical algorithm for variation graph induction that can scale to the largest available pangenomes.

@@

The same encoding can be used to replace the sequence-to-graph mapping $Z$, yielding $\mathcal{Z}$.
Rather than pairs of characters in $Q$ and $\mathcal{N}$, we record runs of matches between them as range matches.
Although in expectation the length of these matches should be strictly less than $k$, due to the interruption of the graph by variation between genomes, this still allows us to reduce the size of $Z$ using runs of matches between $Q$ and $\mathcal{N}$.
Additionally, we store the inverse of $\mathcal{Z}$, which maps ranges from $\mathcal{N} \to Q$, as $\bar{\mathcal{Z}}$.
We use $\mathcal{Z}$ to compact non-branching regions of $\mathcal{V}$ into single nodes, and $\bar{\mathcal{Z}}$ to accelerate our calculation of links in the graph.

@@

To compact $\mathcal{V}$, we traverse $Q$, finding each entry in $\mathcal{Z}$ in turn, recording its start and end in $\mathcal{N}$, which can be understood as a character vector or string containing all the sequence in the nodes of $\mathcal{V}$.
%in a ``graph vector'' $\mathcal{G} = n_1,n_2 \ldots n_{|\mathcal{N}|}$ containing the ordered characters of the graph.
%Any position in $\mathcal{G}$ so marked must be the start or end of a node.
We subsequently use these markings to subdivide $\mathcal{N}$ into a compacted version $\mathcal{N}'$ where compacted node boundaries are marked in a auxiliary bitvector $\mathcal{B} : |\mathcal{B}| = |\mathcal{N}|$ where the first character in each compacted node is marked by a $1$ and other characters are marked $0$.
$\mathcal{B}$ allows us to compute compacted node ids using efficient rank operations \citep{Gog_2014}.
%such that $\forall_{i \in 1 \ldots |\mathcal{N}'|} |n_i| \ge 1$.
%In Algorithm \ref{alg:seqwish} we provide a detailed description of the graph induction algorithm.

@@

We have presented a complete model for variation graph induction from sequences and their pairwise alignments (Algorithm \ref{alg:seqwish}).
Here, we describe our specific implementation of this algorithm: \textit{seqwish}.
In general, our approach uses external memory to elaborate the graph, taking advantage of the availability of low-latency storage media, like solid-state drives (SSDs), to maximize the performance of this approach.

@@

We have presented a straightforward algorithm to generate a pangenome graph from a collection of genomes and alignments between them.
By exploiting a simple model of this algorithm, we provide computational bounds that give insight into the complexity of the problem.
We then make this approach practical by applying the concept of \textit{match compression}, which reduces the expected computational complexity by a factor proportional to the diversity of input sequences.
Our experimental results demonstrate that we can apply our method to various collections of sequences and alignments.
It easily scales to some of the largest species pangenome construction problems possible using publicly-available, high-quality genome assemblies.
%Although we have focused on the problem of building pangenomes,
\textit{seqwish} is a generic sequence graph inducer of potentially many uses.
We envision that it can serve as a component in diverse sequence analysis and assembly pipelines, and hope that our thorough description of its core algorithm and functionality will enable its reuse by other researchers.

@@

\section*{Acknowledgments}
